# Dimensionality reduction techniques

### Data exploration

This time we are using a dataset from the United Nations Development Programme. The dataset called "human" includes several variables from different countries around the world. Below are the explanations of variables.

* SecoEd_FM = the female to male ratio of having at least secondary education
* LaboFor_FM = the female to male ratio of people in the working force
* ExpEdu = expected years of schooling
* LifeE = life expectancy at birth
* GNI = Gross National Income per capita
* MatMort = maternal mortality ratio
* AdoBirth = Adolescent birth rate
* ParlRep = Percetange of female representatives in parliament

Let's explore the data a bit now
```{r data expl}
human <- human <- read.csv(file = "data/human.csv", header = TRUE, sep = ",", row.names = 1)
str(human)
dim(human)
library(dplyr)
``` 

Ok, for some reason GNI is now an integer although I succefully changed it to be numeric....the values still look the same as after I changed it to numeric, so this probably has no effect on anything at all. I will leave it as it is.

```{r overview}
library(ggplot2); library(GGally); library(corrplot)
ggpairs(human)
cor(human)%>% corrplot(type = "upper") 
```

The SecoEd_FM and LaboFor_FM variables have a right-skewed distribution, but does mostly not exceed 1. This means that in most countries more males have at least secondary education than females and more males are in the working force than females. Based on what the relationship between GNI and the two previous variables we can say that the female/male ratios are especially low in poor countries with low GNI. Similar relationship can be seen with e.g. GNI and life expectancy: in poor countries life expectancy is lower than in countries with high GNI. 

If we look at the correlation plot, we can see that the highest positive correlations lie between life expectancy at birth and expected years of education, and adolescent birth and maternal mortality. The highest negative correlation lies between life expectancy at birth and maternal mortality. GNI is has highest (positive) correlation with expected education. 

### Principal Component Analysis (PCA)

Alright, first let's perform PCA on the non-standardized data

```{r PCA non}
pca_human <- prcomp(human)
s <- summary(pca_human)
s
pca_pr <- round(1*s$importance[2, ]*100, digits = 1)
pca_pr
pc_lab<-paste0(names(pca_pr), " (", pca_pr, "%)")
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])
```

And next with the standardized data

```{r PCA standardized}
human_std <- scale(human)
pca_humanstd <- prcomp(human_std)
s <- summary(pca_humanstd)
s
pca_pr <- round(1*s$importance[2, ]*100, digits = 1)
pca_pr
pc_lab<-paste0(names(pca_pr), " (", pca_pr, "%)")
biplot(pca_humanstd, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])
```

Clearly the latter figure is more readable than the previous. There were also some zero-length arrows in the PCA of non-standardized data, which led to them being left out completely. The standardization changed the data so that PCA is able to take into account all data there is. 

The latter PCA shows that there are differences between variables and their relationships to the principal components. This can be seen by looking at the arrows and their angles towards each other and to the PCs. In the first figure it looks like there are no differences at all, as all the arrows have no angle with the PCs. The smaller the angle is, whether between the different arrows, or the arrow and PC axis, the higher positive correlation there is. So we see that female parliament representation and female/male labour force ratio have quite small correlations with all other variables and also with PC1.


### FactomineR and MCA 


```{r factominer}
library(FactoMineR);library(tidyr)
data("tea")
str(tea)
dim
gather(tea) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()
```

The figure is a bit messy, but unfortunately I don't have time now to try to make it clearer. It doesn't work the same as in datacamp, probably because I have all the variables left here instead of choosing a smaller dataset like in the datacamp examples. Next I will, anyway perform mutliple correspondence analysis on the whole dataset. Let's see what that'll look like..

```{r MCA}
library(stringr)
tea$age <- str_replace(tea$age, pattern=",", replace ="") %>% as.factor() #this changes variable age to a factor, since all variables need to be factors
mca <- MCA(tea, graph = FALSE)
plot(mca, invisible=c("ind"), habillage = "quali")
summary(mca)
```



